Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-10 paper_io_sim.tex
edited methods

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, write-through enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends write-through
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform write-through as well as
            direct I/O as it delays disk writes to perform a bulk write at a later time
            \cite{linuxdev3rd2010}.

@@

    \section{Method}
    \label{method}
    We separate our simulation model in two components, the IO
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the IO Controller. The IO Controller interacts as needed with
    the Memory Manager to free memory by triggering flushing or eviction,
    and interact with cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction from cache, and reads or writes to disk when necessary.
    In case the write-through strategy is used, the IO Controller can also interact directly
    with disk model to simulate disk writes.

    \begin{figure}
           \centering
           \includegraphics[width=0.85\columnwidth]{figures/interaction.pdf}
           \caption{Overview of the page cache simulator.
           Applications send file read or write requests to the
           IO Controller that orchestrates flushing, eviction, cache
           and disk accesses with the Memory Manager. Concurrent accesses to storage
           devices (memory and disk) are simulated using existing models.}
           \label{fig:interaction}
    \end{figure}

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by its storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

    In the Linux kernel, page cache LRU lists contain file pages. However,
    due to the large number of file pages, managing lists of pages
    induces substantial overheads.
    Therefore, we introduce the concept of a data block as a unit to represent data
    cached in memory. A data block is a subset of file pages stored in
    page cache that were accessed in the same I/O operation.
    A data block has information about file name, block size, last access
    time, a dirty flag that represents whether the data is clean (0)
    or dirty (1), and an entry (creation) time.
    Blocks can have different sizes and a given file can have multiple
    data blocks in page cache. In addition, a data block can be split into an
    arbitrary number of smaller blocks.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/lru_lists.pdf}
           \caption{Model of page cache LRU lists with data blocks.}    \label{fig:lrulist}
    \end{figure}

@@

    At any given time, a file can be partially cached, completely cached,
    or uncached at all. A cached data block can only reside in one of two
    page cache LRU lists. The first time they are accessed, blocks are
    added to the inactive list. On subsequent accesses, blocks of the
    inactive list are moved to the top of the active list. Cached blocks
    written to cache are marked as dirty until they are flushed to disk.

@@

    \begin{algorithm}\caption{Periodical flushing simulation in IO Controller}\label{alg:pdflush}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{in}{page cache inactive list}
                \Desc{ac}{page cache active list}
                \Desc{t}{predefined flushing time interval}
                \Desc{exp}{predefined expiration time}
                \Desc{sm}{storage simulation model}
               \EndInput
               \While{host is on}
                \State blocks = expired\_blocks(exp, in) + expired\_blocks(exp, ac)
                \State flushing\_time = sm.write(blocks)
                \State for blk in blocks: blk.dirty = 0
                \If{flushing\_time $<$ t}
                    \State sleep(t - flushing\_time)
                \EndIf  % Else: do nothing, continue the loop
            \EndWhile
        \end{algorithmic}
    \end{algorithm}

    Periodical flushing is simulated in the Memory Manager
    background thread. As in the Linux kernel, a dirty block
    in our model is considered expired if (1) it is dirty, and (2)
    the duration since its entry time is longer than a
    predefined expiration time.
    Periodical flushing is simulated as an infinite loop in which
    the Memory Manager searches and flushes these expired dirty blocks to disk (Algorithm~\ref{alg:pdflush}).
    % From Tristan: the algorithm is quite straightforward, I don't think this is
    % necessary
    % In each repetition, Memory Manager finds expired dirty blocks in two
    % page cache LRU lists (line 8), simulates writes of data of these blocks
    % to disk (line 9), and mark them as clean (line 10).
    % If the flushing time does not exceed our time interval, the thread is put
    % to sleep for the remaining time (lines 11-13).
    % Then based on the state of the host at current simulated time,
    % the algorithm continues or finishes the loop.
    Because periodical flushing is simulated as a background thread, it can happen concurrently
    with disk I/O initiated by the main thread. This is taken into account by the
    storage model and reflected in simulated I/O time.

@@

    \begin{algorithm}\caption{File chunk read simulation in IO Controller}
    \label{alg:read}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{fs}{file size (assumed to fit in memory)}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
               \State disk\_read = min(cs, fs - mm.cached(fn)) \Comment{To be read from disk}
               \State cache\_read = cs - disk\_read \Comment{To be read from cache}
               \State required\_mem = cs + disk\_read
               \State mm.flush(cs + disk\_read - mm.free\_mem - mm.evictable, fn)
               \State mm.evict(cs + disk\_read - mm.free\_mem, fn)
               \If {disk\_read $>$ 0}  \Comment{Read uncached data}
               \State sm.read(disk\_read)
               \State mm.add\_to\_cache(disk\_read, fn)
               \EndIf
               \If {cache\_read $>$ 0} \Comment{Read cached}
               \State mm.cache\_read(cache\_read)
            \EndIf
            \State mm.use\_anonymous\_mem(cs)
        \end{algorithmic}
    \end{algorithm}
    As mentioned previously, our model reads and writes file chunks in a
    round-robin fashion. To read a file chunk, simulated applications send
    chunk read requests to the IO Controller which processes them using
    Algorithm~\ref{alg:read}. First, we calculate the amount of uncached
    data that needs to be read from disk, and the remaining amount is read
    from cache (line 6-7). The amount of memory required to read the chunk
    is calculated, corresponding to a copy of the chunk in anonymous memory
    and a copy of the chunk in cache (line 8).
    If there is not enough available memory, the Memory Manager is called
    to flush dirty data (line 9). If necessary, flushing is complemented by
    eviction (line 10). Note that, when called with negative arguments, functions
    \texttt{flush} and \texttt{evict} simply return and do not do anything. Then, if there is
    uncached data, memory manager is called to read data from disk and add this
    data to cache (line 12).
    If cached data needs to be read, the Memory Manager is called to simulate
    a cache read  and update the corresponding data blocks accordingly (line 15).
    Finally, the memory manager is called to reduce the amount of free memory
    corresponding to the amount of anonymous memory used by application (line 17).

    \begin{algorithm}\caption{File chunk write simulation in IO Controller}
    \label{alg:write}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{mm}{MemoryManager object}
               \EndInput
            \State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
            \If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
                \State mm.evict(min(cs, remain\_dirty) - mm.free\_mem)
                \State mem\_amt = min(cs, mm.free\_mem)
                \State mm.write\_to\_cache(fn, mem\_amt)
            \EndIf
            \State remaining = cs - mem\_amt
            \While {remaining $>$ 0}  \Comment{Write with disk bandwidth}
                \State mm.flush(cs - mem\_amt)
                \State mm.evict(cs - mem\_amt  - mm.free\_mem)
                \State to\_cache = min(remaining, mm.free\_mem)
                \State mm.write\_to\_cache(fn, to\_cache)
                \State remaining = remaining - to\_cache
            \EndWhile

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes our simulation of chunk writes in
    the IO Controller.
    % Data is written to cache until the amount of dirty data in the cache
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    Our algorithm initially checks the  amount of data that
    can be written as dirty data given the dirty ratio (line 5).
    If this amount is greater than 0, the Memory Manager is requested to evict
    data from cache if necessary (line 7).
    After eviction, the amount of data that can be written to
    page cache is calculated (line 8), and a cache write is simulated (line 9).
    If there is remaining data, this data is written to cache in a loop (line 12 - line 18).
    In this loop, we flush and evict data as possible so that there will be
    available space for writing (line 13-14).
    After flushing and eviction, the amount of data that can be written
    is calculated (line 15) and a cache write is simulated (line 16).
    We update the remaining amount (line 17) and continue the loop until all data
    is written.

    In addition to the chunk write simulation with writeback page cache above,
    we also add a chunk write function with write-through strategy to the IOController.
    This function simply simulates a disk write with the amount of data passed in,
    then evicts cache if needed and adds this data to cache.

@@

            To validate our simulation model, we created a simple prototype
            simulator independent of existing simulation frameworks and libraries.
            This allowed us to evaluate the accuracy and correctness of our
            model in a simple scenario before integrating it to the more complex
            WRENCH framework.
            In this prototype we used the following basic storage model for
            both memory and disk:
            \begin{align*}
                & t_{r} = D / b_r \\
                & t_{w} = D / b_w\
            \end{align*}

@@

            Since bandwidth sharing was not simulated, this prototype does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}. We used Python 3.7.

            We also implemented our model as part of WRENCH, enhancing its internal implementation
            and APIs with a page cache abstraction, and allowing users to activate the feature via a
            command-line argument. For the experiments in this paper we use WRENCH 1.6 at this commit \url{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}, which uses
            SimGrid 3.25 (available at \url{https://framagit.org/simgrid/simgrid}). However, our implementation is now part of WRENCH's master branch and will be available to users with the next WRENCH release.

\tristan{anything else interesting to mention about the implementation?}
\henri{I mentioned that the implementaion is now part of the WRENCH master }

@@

        Our experiments compare real executions with our Python prototype,
        with the original WRENCH, and with our WRENCH-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems.

        We simulate an application that consists of three single-core, sequential tasks
        where each task reads the file produced by the previous task,
        increments every byte of this file to emulate real processing, and
        writes the resulting data to disk.
        In real execution, we clear cache before starting the application to make sure
        the conditions are consistent, the cache is not in use and it is only used by
        our application.
        In addition, the anonymous memory used by application
        is released after each task in the pipeline.
        Thus, we also release the memory used at the end of each task in our simulator.
        We also use a real application
        \textcolor{red}{[Maybe briefly describe this application's purpose
        (e.g. a standard fMRI preprocessing piepeline)]}, to evaluate the
        applicability of our simulation model.

        The real application runs on a dedicated cluster hosted at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a network bandwidth of 25 Gbps. Each
        compute node has 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 275~GB (256~GiB) \tristan{are you sure it's not 256GB?
        pls check.} of RAM, 6 $\times$ SSDs of 450~GB each with the XFS
        file system, 378~GB of tmpfs, and 126~GB of devtmpfs file system.
        Nodes run CentOS~8.1 with NFS version 4. We use the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput.

        To parametrize our simulators, we benchmarked the bandwidths of
        memory, local disk, remote disk (NFS), and network
        (Table~\ref{table:benchmark}). Since SimGrid, and thus WRENCH, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}[htbp]
            \centering
            \begin{tabularx}{\columnwidth}{|l
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X|}
            \hline
                Bandwidths  & Cluster (real) & Python simulator & WRENCH simulator\\
            \hline
                Memory read  & 6860    & 4812     & 4812\\
                Memory write & 2764    & 4812 & 4812\\
                Local disk read & 510 & 465 & 465\\
                Local disk write & 420 & 465     & 465\\
                Remote disk read & 515 & - & 445\\
                Remote disk write & 375 & - & 445\\
                Network & 3000 & - & 3000\\
            \hline
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator. \tristan{table could be prettified}}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task durations on a cluster node and used these
            durations in our simulation. For the Python simulator,
            execution times were directly used in the simulation. For
            WRENCH and WRENCH-cache, we determined the corresponding number
            of flops on a 1~Gflops CPU and used these values in the
            simulation (see Table~\ref{table:cputime}). The final simulated
            platform and application used in our experiment are avaiable at
            \url{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}.

            \begin{table}[htbp]
            \centering
            \begin{tabularx}{0.8\columnwidth}{|l|>{\centering\arraybackslash}X|}
            \hline
                Input size (GB)  & CPU time (second)\\
            \hline
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \hline
            \end{tabularx}
            \caption{Task CPU time corresponding to input file size.}
            \label{table:cputime}
            \end{table}

@@

            Our first experiment (\textit{Exp single-threaded}) simply runs one instance of the
            application on a single cluster node, with different input file
            sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed
            to the same local disk. The goal is to validate the simulation model
            in absence of any concurrent disk access.

            Our second experiment (\textit{Exp concurrent}) runs concurrent instances of the
            application on a single node, all application instances
            operating on different files of size 3~GB stored in the same
            local disk, sharing the disk bandwidth. We vary the number of
            concurrent application instances from 1 to 32 since cluster
            nodes have 32 CPU cores.

            Our third experiment (\textit{Exp NFS}) occurs in the same configuration as the
            previous one, albeit reading and writing on a 400-GB NFS-mounted remote disk
            mounted from another compute node. As is commonly
            configured in HPC environments to avoid data loss, in our simulator,
            NFS client and server read caches are enabled, but there is no cache
            for writing on client side and server writes are configured with
            write-through cache instead of writeback.
            Therefore, all the writes happen at disk bandwidth, but
            reads can benefit from cache hits. We simulate write-through by
            writing data directly to disk and adding that data to the server
            cache.

            Finally, we simulate a real application from the neuroimaging
            domain. \textcolor{red}{[Description of the real pipeline with
            nighres (including the workflow engine if it's Dask)]}.
            We simulate this pipeline with simulators using original WRENCH
            and WRENCH-cache. Because our work focuses on I/O time, we
            assume that CPU time is correctly modeled and use the CPU time
            measured in the real pipelines to setup our simulated
            application.

@@

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, owing to the fact
        that all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, part
        of it is on disk. This is due to the fact that the simulated flushing and eviction
        was triggered after File 3 was already written, while in reality these operations
        happened before File 3 was written \tristan{Dzung, pls check that. Also, why doesn't it happen irl?}.

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only write-through cache was
            enabled in the NFS server. Both WRENCH and WRENCH-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits \tristan{Dzung, pls check that.}.

@@

        \tristan{discussion could be fleshed out}

>>>>

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, writethrough enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends writethrough
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform writethrough as well as
            direct I/O as it delays disk writes to perform a bulk write at a later time
            \cite{linuxdev3rd2010}.

@@

    \section{Methods}
    \label{method}
    We separate our simulation model in two components, the I/O
    Controller and the Memory Manager, which together simulate
    file reads and writes (Figure~\ref{fig:interaction}).
    To read or write a file chunk, a simulated application sends a
    request to the I/O Controller. The I/O Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction, and reads or writes to disk when necessary.
    In case the writethrough strategy is used, the I/O Controller directly
    writes to disk \tristan{Then how is the written data added to cache?}.

    \begin{figure}
           \centering
           \includegraphics[width=0.85\columnwidth]{figures/interaction.pdf}
           \caption{Overview of the page cache simulator.
           Applications send file read or write requests to the
           I/O Controller that orchestrates flushing, eviction, cache
           and disk accesses with the Memory Manager. Concurrent accesses to storage
           devices (memory and disk) are simulated using existing models.}
           \label{fig:interaction}
    \end{figure}

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by their storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

    In the Linux kernel, page cache LRU lists contain file pages. However,
    due to the large number of file pages, simulating lists of pages
    induces substantial overheads.
    Therefore, we introduce the concept of a data block as a unit to represent data
    cached in memory. A data block is a subset of file pages stored in
    page cache that were accessed in the same I/O operation.
    A data block has information about the file name, block size, last access
    time, a dirty flag that represents whether the data is clean (0)
    or dirty (1), and an entry (creation) time.
    Blocks can have different sizes and a given file can have multiple
    data blocks in page cache. In addition, a data block can be split into an
    arbitrary number of smaller blocks.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/lru_lists.pdf}
           \caption{Model of page cache LRU lists with data blocks.}
           \label{fig:lrulist}
    \end{figure}

@@

    At any given time, a file can be partially cached, completely cached,
    or uncached at all. A cached data block can only reside in one of the two
    LRU lists. The first time they are accessed, blocks are
    added to the inactive list. On subsequent accesses, blocks of the
    inactive list are moved to the top of the active list. Cached blocks
    written to cache are marked as dirty until they are flushed to disk.

@@

    \begin{algorithm}\caption{Periodical flushing simulation in I/O Controller}\label{alg:pdflush}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{in}{page cache inactive list}
                \Desc{ac}{page cache active list}
                \Desc{t}{predefined flushing time interval}
                \Desc{exp}{predefined expiration time}
                \Desc{sm}{storage simulation model}
               \EndInput
               \While{host is on}
                \State blocks = expired\_blocks(exp, in) + expired\_blocks(exp, ac)
                \State flushing\_time = sm.write(blocks)
                \State for blk in blocks: blk.dirty = 0 \tristan{with this logic a flushed block can remain dirty for a long time. Is it what you implemented?
                it would be wise to set the dirty flag to 0 right after the block is written}
                \If{flushing\_time $<$ t}
                    \State sleep(t - flushing\_time)
                \EndIf  % Else: do nothing, continue the loop
            \EndWhile
        \end{algorithmic}
    \end{algorithm}

    Periodical flushing is simulated in the Memory Manager
    background thread. As in the Linux kernel, a dirty block
    in our model is considered expired if
    the duration since its entry time is longer than a
    predefined expiration time.
    Periodical flushing is simulated as an infinite loop in which
    the Memory Manager searches for dirty blocks and flushes them to disk (Algorithm~\ref{alg:pdflush}).
    % From Tristan: the algorithm is quite straightforward, I don't think this is
    % necessary
    % In each repetition, Memory Manager finds expired dirty blocks in two
    % page cache LRU lists (line 8), simulates writes of data of these blocks
    % to disk (line 9), and mark them as clean (line 10).
    % If the flushing time does not exceed our time interval, the thread is put
    % to sleep for the remaining time (lines 11-13).
    % Then based on the state of the host at current simulated time,
    % the algorithm continues or finishes the loop.
    Because periodical flushing is simulated as a background thread, it can happen concurrently
    with disk I/O initiated by the main thread. This is taken into account by the
    storage model and reflected in simulated I/O time.

@@

    \begin{algorithm}\caption{File chunk read simulation in the I/O Controller}
    \label{alg:read}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{fs}{file size (assumed to fit in memory)}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
               \State disk\_read = min(cs, fs - mm.cached(fn)) \Comment{To be read from disk}
               \State cache\_read = cs - disk\_read \Comment{To be read from cache}
               \State required\_mem = cs + disk\_read
               \State mm.flush(required\_mem - mm.free\_mem - mm.evictable, fn)
               \State mm.evict(required\_mem - mm.free\_mem, fn)
               \If {disk\_read $>$ 0}  \Comment{Read uncached data}
               \State sm.read(disk\_read)
               \State mm.add\_to\_cache(disk\_read, fn)
               \EndIf
               \If {cache\_read $>$ 0} \Comment{Read cached}
               \State mm.cache\_read(cache\_read)
            \EndIf
            \State mm.use\_anonymous\_mem(cs)
        \end{algorithmic}
    \end{algorithm}
    As mentioned previously, our model reads and writes file chunks in a
    round-robin fashion. To read a file chunk, simulated applications send
    chunk read requests to the I/O Controller which processes them using
    Algorithm~\ref{alg:read}. First, we calculate the amount of uncached
    data that needs to be read from disk, and the remaining amount is read
    from cache (line 7-8). The amount of memory required to read the chunk
    is calculated, corresponding to a copy of the chunk in anonymous memory
    and a copy of the chunk in cache (line 9).
    If there is not enough available memory, the Memory Manager is called
    to flush dirty data (line 10). If necessary, flushing is complemented by
    eviction (line 11). Note that, when called with negative arguments, functions
    \texttt{flush} and \texttt{evict} simply return and do not do anything. Then,
    if the block requires
    uncached data, the memory manager is called to read data from disk and to add this
    data to cache (line 14).
    If cached data needs to be read, the Memory Manager is called to simulate
    a cache read  and update the corresponding data blocks accordingly (line 17).
    Finally, the memory manager is called to reduce the amount of free memory
    corresponding to the amount of anonymous memory used by application (line 19).

    \begin{algorithm}\caption{File chunk write simulation in the I/O Controller}
    \label{alg:write}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{mm}{MemoryManager object}
               \EndInput
            \State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
            \If {remain\_dirty $>$ 0} \Comment{Write to memory}
                \State mm.evict(min(cs, remain\_dirty) - mm.free\_mem)
                \State mem\_amt = min(cs, mm.free\_mem)
                \State mm.write\_to\_cache(fn, mem\_amt)
            \EndIf
            \State remaining = cs - mem\_amt
            \While {remaining $>$ 0}  \Comment{Flush to disk, then write to cache}
                \State mm.flush(cs - mem\_amt)
                \State mm.evict(cs - mem\_amt  - mm.free\_mem)
                \State to\_cache = min(remaining, mm.free\_mem)
                \State mm.write\_to\_cache(fn, to\_cache)
                \State remaining = remaining - to\_cache
            \EndWhile

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes our simulation of chunk writes in
    the I/O Controller.
    % Data is written to cache until the amount of dirty data in the cache
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    Our algorithm initially checks the  amount of dirty data that
    can be written given the dirty ratio (line 5).
    If this amount is greater than 0, the Memory Manager is requested to evict
    data from cache if necessary (line 7).
    After eviction, the amount of data that can be written to
    page cache is calculated (line 8), and a cache write is simulated (line 9).
    If the dirty threshold is reached and there is remaining data to be written,
    the remaining data is written to cache in a loop
    where we repeatedly flush and evict to free space in the cache (line 12-18).

    The write simulation described above models page cache in writeback
    mode. Our model also includes a write function in writethrough mode,
    which simply simulates a disk write with the amount of data passed in,
    then evicts cache if needed and adds the written data to the cache.

@@

            We first created a standalone prototype
            simulator to evaluate the accuracy and correctness of our
            model in a simple scenario before integrating it in the more complex
            WRENCH framework.
            The prototype uses the following basic storage model for
            both memory and disk:
            \begin{align*}
                & t_{r} = D / b_r \\
                & t_{w} = D / b_w\
            \end{align*}

@@

            Since bandwidth sharing is not simulated, the prototype does not support
            concurrency: it is limited to single-threaded applications running on systems
            with a single-core CPU. We used this prototype for a first validation of our simulation
            model against a real sequential application running on a real system.
            The source code this prototype is available at
            \url{https://github.com/big-data-lab-team/paper-io-simulation/tree/master/exp/pysim}. We used Python 3.7.

            We also implemented our model as part of WRENCH, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. For the experiments in this paper we use WRENCH 1.6
            at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses SimGrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of WRENCH's master branch and will be available to
            users with the next WRENCH release. WRENCH provides a full SimGrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.

@@

        Our experiments compare real executions with our Python prototype,
        with the original WRENCH, and with our WRENCH-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems. We use two applications: a synthetic one created to
        evaluate the simulation model, and a real one representative of
        neuroimaging data processing.

        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. In real
        executions, we clear the page cache before starting the application,
        to ensure comparable conditions among the runs.
        The anonymous memory used by application
        is released after each task, which we also simulate in the Python prototype and in WRENCH-cache.

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres} that implements skull
        stripping in high-resolution brain images, an operation common to
        many analysis pipelines. The workflow is implemented as a Python
        script that calls Java image-processing routines. We used Python
        3.6, Java 8, and Nighres \tristan{add version}. We patched the
        skull stripping workflow to monitor task CPU times and inject them
        in the simulation, to focus our analysis on I/O times \tristan{add
        link to patched code}.

        The applications were executed on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a network bandwidth of 25 Gbps. Each
        compute node has 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, and 126~GiB of devtmpfs file system.
        Nodes run CentOS~8.1 with NFS version 4. We use the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput.

        To parametrize our simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since SimGrid, and thus WRENCH, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}[htbp]
            \centering
            \begin{tabularx}{\columnwidth}{|l
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X|}
            \hline
                Bandwidths  & Cluster (real) & Python simulator & WRENCH simulator\\
            \hline
                Memory read  & 6860    & 4812     & 4812\\
                Memory write & 2764    & 4812 & 4812\\
                Local disk read & 510 & 465 & 465\\
                Local disk write & 420 & 465     & 465\\
                Remote disk read & 515 & - & 445\\
                Remote disk write & 375 & - & 445\\
                Network & 3000 & - & 3000\\
            \hline
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator. \tristan{table could be prettified}}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python simulator, we used CPU times
            directly in the simulation. For WRENCH and WRENCH-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are avaiable at
            \url{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}.

            \begin{table}[htbp]
            \centering
            \begin{tabularx}{0.8\columnwidth}{|l|>{\centering\arraybackslash}X|}
            \hline
                Input size (GB)  & CPU time (second)\\
            \hline
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \hline
            \end{tabularx}
            \caption{Task CPU time corresponding to input file size. \tristan{are the input files in GB or in GiB? If GiB, then change the abbreviation everywhere.}
            \tristan{add nighres CPU times when available}}
            \label{table:cputime}
            \end{table}

@@

            Our first experiment (\textit{Exp single-threaded}) runs a
            single instance of the synthetic application on a single
            cluster node, with different input file sizes (20~GB, 50~GB,
            75~GB, 100~GB), and with all I/Os directed to the same local
            disk. The goal is to validate the simulation model in absence
            of any concurrent disk access.

            Our second experiment (\textit{Exp concurrent}) runs concurrent instances of the
            application on a single node, all application instances
            operating on different files of size 3~GB stored in the same
            local disk. We vary the number of
            concurrent application instances from 1 to 32 since cluster
            nodes have 32 CPU cores.

            Our third experiment (\textit{Exp NFS}) occurs in the same
            configuration as the previous one, albeit reading and writing
            on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
            another compute node. As is commonly configured in HPC
            environments to avoid data loss, there is no client write cache
            and the server cache is configured as writethrough instead of
            writeback. NFS client and server read caches are enabled. 
            Therefore, all the writes happen at disk bandwidth, but
            reads can benefit from cache hits.

            Our fourth and final experiment (\textit{Exp Nighres})
            simulates the Nighres skull stripping workflow
            \tristan{describe data, and if the execution was single node or
            distributed.}

@@

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, part
        of it is on disk. This is due to the fact that the simulated flushing and eviction
        was triggered after File 3 was already written, while in reality these operations
        happened before File 3 was written \tristan{Dzung, pls check that. Also, why doesn't it happen irl?}.

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only writethrough cache was
            enabled in the NFS server. Both WRENCH and WRENCH-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits \tristan{Dzung, pls check that.}.

@@

        \tristan{discussion could be fleshed out. cgroups}
